---
title: "Week 3 Exercises Solutions"
author: "J.Blondeau"
date: "2024-09-05"
output: 
  html_document:
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

All the following question use the sttst_all_sites.xlsx file in the /data directory.

All answers should be made in the .Rmd file, showing the code and your answer and everything pushed back up to your GitHub repo.

triple stars renders a thin line across the page

***


Import libraries but don't show the output 'include=FALSE'
```{r results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(here) ### this library helps you with filepaths
```

1. Import sttstj_all_sites and inspect the data set.  What is the data structure and the data type of each field?


```{r}
##this is a code chunk, anything inside the back tics is like being in the R console.

## Inspect data
data_location <- here("data/sttstj_all_sites.xlsx")
print(data_location)

##Inspect the excel sheets available
excel_sheets(data_location)
read_xlsx(data_location, sheet = "Metadata")

## Read in data
data <- read_xlsx(data_location ,sheet = "locations")
class(data)
glimpse(data)
```
***

2. How many samples are in the data set?   

```{r}
print(nrow(data))
count(data)

## What is this code doing? Are there other ways to accomplish this same task?
length(unique(data$yr_site))
```


3. How many unique habitat types are there?  What are the names of each habitat?

```{r}
hab_types <- unique(data$hab)
print(hab_types)
```


4. How many sites are there in the "PVMT" habitat type?

```{r}
data %>% 
  count(hab)

```


5. How many sites are there in the "PVMT" or "AGRF" habitat types? You could manually sum the site numbers using previous code, but how would you do this programmatically?

```{r}
data %>% 
  filter(hab %in% c("PVMT","AGRF")) %>% 
  count(hab) %>% 
  summarise(tot = sum(n))

```


6. How many sites are there in all the habitat types EXCEPT "PVMT"?

```{r}
data %>% 
  filter(hab != "PVMT") %>% 
  count(hab) %>% 
  summarise(tot = sum(n))
```


7. Show the first 5 records (rows) of...
    - all sites sampled in "AGRF" or "SCR"
    - only show the "yr_site" and "hab" columns
    
```{r}
data %>% 
  filter(hab %in% c("AGRF", "SCR")) %>% 
  head()

data %>% 
  filter(hab %in% c("AGRF", "SCR")) %>% 
  select(yr_site, hab) %>% 
  head()
```

    
<!-- extra line break -->
<br>

8. What is the northern most site sampled in the "BDRK" habitat?

```{r}
data %>% 
  filter(hab == "BDRK") %>% 
  arrange(desc(lat))
```


9. What is the western most site sampled in either "AGRF" or "PTRF" habitat?

```{r}
data %>% 
  filter(hab == "AGRF" | hab == "PTRF") %>% 
  arrange(lon)
  
```


10. How many years were surveyed in the data set?

```{r}
data %>% 
  mutate(year = as.integer(substr(yr_site, 1,4))) %>% 
  summarise(length(unique(year)))
```


11. What is the southern most site sample in the "PVMT" habitat in 2004?

```{r}
data %>% 
  mutate(year = as.integer(substr(yr_site, 1,4))) %>% 
  filter(year == 2004 & hab == "PVMT") %>% 
  arrange(lat)
```




